---
title: "When Do Random Effects Models Reduce to
  Fixed-Effect Alternatives? Conditions for
  Equivalence in Longitudinal Clinical Trials"
author: "R.G. Thomas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
header-includes:
   - \usepackage{placeins}
   - \usepackage{amsmath}
   - \usepackage{bm}
   - \rhead{Lab report}
   - \fancyfoot[L]{\currfilename}
   - \setlength{\headheight}{13.59999pt}
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

```{r setup, include=FALSE, echo=FALSE}
rm(list = ls())
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9,
  knitr.table.format = "latex"
)
library(knitr)
library(kableExtra)
library(tidyverse)
library(nlme)
library(lme4)
library(emmeans)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  results = "asis",
  dev = "pdf",
  cache = TRUE,
  cache.path = "cache/"
)
options(scipen = 1, digits = 4)
```

# Introduction

## The practical problem

In clinical trials with repeated measurements, mixed
models for repeated measures (MMRM) and random
effects models have become the standard analytic
approach [@mallinckrodt2001type; @mallinckrodt2008recommendations].
These models accommodate within-subject correlation,
handle incomplete data under the missing-at-random
assumption, and provide flexible covariance
structures. However, they impose a substantial
interpretive burden on collaborators unfamiliar with
the statistical machinery of variance components,
restricted maximum likelihood estimation, and
covariance parameterization.

A natural question arises: under what conditions does
the random effects model reduce to, or closely
approximate, a simpler fixed-effect analysis that
would be more accessible to the statistical
practitioner? If such conditions can be identified,
the simpler model offers an intuitive bridge to the
same inferences, sparing the analyst from invoking
a complex apparatus when it is not strictly
necessary.

## Foundational results

The relationship between random effects and
fixed-effect analyses has been explored from
several angles. @brogan1980comparative showed that
in a two-group pretest-posttest design, the
repeated-measures ANOVA subsumes the difference-score
$t$-test: the interaction $F$-statistic equals the
square of the two-sample $t$-test on difference
scores. The difference-score analysis requires fewer
assumptions (homogeneous variance of differences
and normally distributed errors) than the full
repeated-measures model (equal variance-covariance
matrices across groups). This algebraic embedding
demonstrates that when the research question
concerns only whether treatment groups differ in
their pre-to-post change, the simpler analysis is
not merely an approximation but is arithmetically
identical to a component of the more elaborate
model.

@laird1982random introduced the general framework
for random-effects models in longitudinal data that
has since become ubiquitous. In their formulation,
the random effects induce a structured
variance-covariance matrix on the marginal
distribution of the observed data. The key insight
for our purposes is that when this marginal
covariance is left unconstrained---as in the
unstructured covariance MMRM---the random effects
become a computational device rather than a
substantive modeling choice.

## The Frost-Kenward-Fox observation

@frost2008optimizing established a methodological
framework for comparing conventional and run-in
trial designs using linear mixed models. In their
treatment of the two-timepoint case, they showed
that the Laird-Ware random-intercept formulation
treating baseline and follow-up as correlated
outcomes (their equation 8) yields identical point
estimates and very similar standard errors to the
analysis of covariance model that conditions on
baseline (their equation 6). This near-equivalence
holds provided the variance-covariance matrix is
unconstrained and $\sigma_{12} < \sigma_1^2$ and
$\sigma_1^2 = \sigma_2^2$ (i.e., homogeneous
marginal variances and positive but not perfect
correlation). The result builds on @lee1974note, who
showed that Rao's reduction of the @potthoff1964generalized
generalized MANOVA model yields estimators
identical to those from an analysis of covariance.

## Constrained longitudinal data analysis

@liang2000longitudinal formalized the constrained
longitudinal data analysis (cLDA), in which
baseline is included in the response vector and the
model constrains the two treatment groups to share a
common baseline mean---a consequence of
randomization. @liu2009baseline compared cLDA to
the longitudinal ANCOVA model and showed that (a)
with complete data, the treatment effect estimates
are effectively identical but the cLDA standard
errors are marginally smaller; and (b) with
missing baseline data, cLDA is more efficient
because it can borrow information from the observed
outcome model to impute baseline values implicitly
under the missing-at-random assumption.
@lu2009sample derived sample-size formulae for
cLDA, and @lu2010efficiency provided a formal
efficiency comparison of cLDA versus longitudinal
ANCOVA, quantifying the power gain as a function of
the proportion of missing baselines.

## The role of summary statistics

An alternative thread, developed by
@frison1992repeated and @frison1997linearly,
advocates analyzing repeated measures through
summary statistics---typically the mean post-baseline
value or a fitted slope for each subject---followed
by a between-group comparison of these summaries.
@senn2000repeated extended this approach, showing
that summary-statistic methods can be viewed as
special cases of the general mixed model when the
within-subject covariance is compound symmetric.
@senn2006change provided an influential
clarification of the debate between change-score
analysis and ANCOVA, demonstrating that ANCOVA does
not require baseline balance for unbiasedness and
that the common slope assumption is testable and
often reasonable.

## Modern MMRM practice

In contemporary pharmaceutical practice, MMRM with
unstructured covariance has become the de facto
primary analysis for continuous outcomes
[@mallinckrodt2001type; @mallinckrodt2008recommendations].
@sullivan2022mmrm showed that MMRM achieves power
gains over complete-case ANCOVA only when
time-by-covariate interactions are included;
without these interactions, ANCOVA can be more
powerful. @wang2024improving proposed augmented
MMRM estimators that improve robustness against
model misspecification. @kenward1997small
provided the small-sample degrees-of-freedom
correction that makes MMRM valid in trials with
modest sample sizes, which is essential for the
settings where the random-effects and fixed-effect
models are most likely to be practically
interchangeable.

Textbook treatments of these relationships can be
found in @fitzmaurice2011applied,
@verbeke2000linear, and @diggle2002analysis.
@coffman2016condition provided a recent tutorial
contrasting conditional and marginal approaches to
analyzing change in randomized trials.

## Present study

The present report investigates, through analytic
argument and Monte Carlo simulation, the conditions
under which random effects models for two-group
longitudinal designs reduce exactly or approximately
to fixed-effect alternatives. We consider three
canonical settings:

1. **Two-timepoint (pre-post) designs:** MMRM with
   unstructured $2 \times 2$ covariance versus ANCOVA
   with baseline as covariate.
2. **Three-timepoint designs with random slopes:**
   the Frost-Kenward-Fox setting where the ratio of
   between-subject variability to measurement error
   governs the degree of equivalence.
3. **Constrained longitudinal data analysis versus
   ANCOVA:** the Liang-Zeger formulation and its
   relationship to the conditional model.

For each setting, we identify the parameter
configurations under which the simpler model
provides equivalent or near-equivalent inference,
and we quantify the discrepancy when conditions are
not met.

\FloatBarrier

# Methods

## Setting 1: two-timepoint pre-post design

Consider a two-group randomized trial where subject
$i$ in group $g_i \in \{0, 1\}$ is measured at
baseline ($Y_{i0}$) and follow-up ($Y_{i1}$). The
data-generating model is:

$$
\begin{pmatrix} Y_{i0} \\ Y_{i1} \end{pmatrix}
\sim \mathrm{N}\left[
\begin{pmatrix} \alpha_0 \\ \alpha_1 + \gamma g_i
\end{pmatrix},
\begin{pmatrix} \sigma_1^2 & \sigma_{12} \\
\sigma_{12} & \sigma_2^2
\end{pmatrix}
\right]
$$

where $\gamma$ is the treatment effect at follow-up.
Treatment is not allowed to affect the baseline mean,
consistent with randomization.

**ANCOVA model.** Regress $Y_{i1}$ on $g_i$ and
$Y_{i0}$:

$$
Y_{i1} | Y_{i0} \sim
\mathrm{N}(\alpha + \gamma g_i + \phi Y_{i0},
\; \sigma^2)
$$

**MMRM model.** Fit both timepoints as outcomes with
an unstructured $2 \times 2$ covariance:

$$
\begin{pmatrix} Y_{i0} \\ Y_{i1} \end{pmatrix}
\sim \mathrm{N}\left[
\begin{pmatrix} \mu_0 \\ \mu_1 + \gamma g_i
\end{pmatrix},
\bm{\Sigma}
\right]
$$

We verify algebraically and via simulation that
$\hat{\gamma}$ and its standard error are identical
across the two models when $\bm{\Sigma}$ is
unconstrained and data are complete.

## Setting 2: three-timepoint random slopes

Following @frost2008optimizing, consider three
evenly spaced measurements at $t = 0, t, 2t$ with a
random intercept and slope model:

$$
Y_{ij} | a_i, b_i \sim
\mathrm{N}[\alpha + (\beta + \gamma g_i) t_j +
a_i + b_i t_j, \; \sigma^2]
$$

where $(a_i, b_i)' \sim \mathrm{N}(\mathbf{0},
\mathbf{G})$ with $\mathbf{G}$ containing the
variance of intercepts ($\sigma_a^2$), slopes
($\sigma_b^2$), and their covariance ($\sigma_{ab}$).

The corresponding fixed-effect analysis considers
only the change scores $C_{ij} = Y_{ij} - Y_{i(j-1)}$
and compares groups via a two-sample test. The
degree of equivalence depends on the ratio
$\sigma_b / \sigma$: when between-subject
variability in slopes is large relative to
measurement error, the random effects model gains
appreciable efficiency; when this ratio is small,
the two approaches converge.

## Setting 3: cLDA versus ANCOVA

Following @liang2000longitudinal and
@liu2009baseline, we compare:

- **cLDA:** baseline and follow-up as joint outcomes,
  with the constraint that both treatment groups
  share a common baseline mean.
- **ANCOVA:** follow-up as outcome, baseline as
  covariate.

We evaluate the two approaches under (a) complete
data and (b) 10--30\% missing baselines (MCAR and
MAR mechanisms).

## Simulation procedure

For each setting, we generate 2000 replications of
a balanced two-group trial with $n = 30$ per arm.
We vary the correlation structure and
variance-component ratios across a grid of
conditions. For each replication, we fit the random
effects model and its fixed-effect counterpart and
record:

- Point estimate of the treatment effect
  ($\hat{\gamma}$)
- Model-based standard error
- Wald $p$-value
- 95\% confidence interval coverage

```{r sim-functions, echo=FALSE, cache=TRUE}
sim_prepost <- function(n_per_arm, gamma_true,
                        sigma1, sigma2, rho,
                        n_sims = 2000,
                        seed = 42) {
  set.seed(seed)
  sigma12 <- rho * sigma1 * sigma2
  Sigma <- matrix(
    c(sigma1^2, sigma12, sigma12, sigma2^2),
    nrow = 2
  )
  L <- chol(Sigma)

  results <- tibble(
    sim = integer(),
    method = character(),
    est = numeric(),
    se = numeric(),
    pval = numeric(),
    ci_lo = numeric(),
    ci_hi = numeric()
  )

  for (s in seq_len(n_sims)) {
    n <- 2 * n_per_arm
    g <- rep(0:1, each = n_per_arm)
    Z <- matrix(rnorm(n * 2), ncol = 2) %*% L
    y0 <- 50 + Z[, 1]
    y1 <- 50 + gamma_true * g + Z[, 2]

    dat <- tibble(
      id = factor(rep(seq_len(n), 2)),
      group = factor(rep(g, 2)),
      time = factor(rep(c("pre", "post"),
        each = n)),
      y = c(y0, y1)
    )
    dat_wide <- tibble(
      id = factor(seq_len(n)),
      group = factor(g),
      y0 = y0,
      y1 = y1
    )

    fit_ancova <- lm(y1 ~ group + y0,
      data = dat_wide)
    cf_a <- summary(fit_ancova)$coefficients
    est_a <- cf_a["group1", "Estimate"]
    se_a <- cf_a["group1", "Std. Error"]
    pval_a <- cf_a["group1", "Pr(>|t|)"]

    fit_mmrm <- tryCatch({
      gls(y ~ time * group,
        data = dat,
        correlation = corSymm(form = ~ 1 | id),
        weights = varIdent(form = ~ 1 | time))
    }, error = function(e) NULL)

    if (!is.null(fit_mmrm)) {
      em <- emmeans(fit_mmrm,
        pairwise ~ group | time)
      con <- summary(em$contrasts)
      post_row <- con[con$time == "post", ]
      est_m <- -post_row$estimate
      se_m <- post_row$SE
      pval_m <- post_row$p.value
    } else {
      est_m <- NA; se_m <- NA; pval_m <- NA
    }

    results <- bind_rows(results, tibble(
      sim = s,
      method = c("ANCOVA", "MMRM"),
      est = c(est_a, est_m),
      se = c(se_a, se_m),
      pval = c(pval_a, pval_m),
      ci_lo = c(est_a - 1.96 * se_a,
        est_m - 1.96 * se_m),
      ci_hi = c(est_a + 1.96 * se_a,
        est_m + 1.96 * se_m)
    ))
  }
  results
}
```

```{r sim-run-prepost, cache=TRUE, dependson='sim-functions'}
scenarios <- expand_grid(
  rho = c(0.3, 0.6, 0.9),
  sigma_ratio = c(0.5, 1.0, 2.0)
)

sim_results <- pmap_dfr(scenarios, function(rho,
                                             sigma_ratio) {
  res <- sim_prepost(
    n_per_arm = 30,
    gamma_true = 3,
    sigma1 = 10,
    sigma2 = 10 * sigma_ratio,
    rho = rho,
    n_sims = 2000
  )
  res |>
    mutate(
      rho = rho,
      sigma_ratio = sigma_ratio
    )
})
```

\FloatBarrier

# Results

## Setting 1: two-timepoint equivalence

```{r results-prepost, cache=TRUE, dependson='sim-run-prepost'}
summary_prepost <- sim_results |>
  filter(!is.na(est)) |>
  group_by(rho, sigma_ratio, method) |>
  summarise(
    mean_est = mean(est),
    mean_se = mean(se),
    emp_se = sd(est),
    power = mean(pval < 0.05),
    coverage = mean(
      ci_lo <= 3 & ci_hi >= 3
    ),
    .groups = "drop"
  )
```

```{r table-prepost}
tab <- summary_prepost |>
  mutate(across(
    c(mean_est, mean_se, emp_se),
    ~ round(., 3)
  )) |>
  mutate(across(
    c(power, coverage),
    ~ round(., 3)
  )) |>
  select(
    `$\\rho$` = rho,
    `$\\sigma_2/\\sigma_1$` = sigma_ratio,
    Method = method,
    `Mean $\\hat{\\gamma}$` = mean_est,
    `Mean SE` = mean_se,
    `Empirical SE` = emp_se,
    Power = power,
    Coverage = coverage
  )

kable(tab,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Comparison of ANCOVA and MMRM estimators",
    "in two-timepoint designs ($n = 30$ per arm,",
    "$\\gamma = 3$, 2000 replications)."
  ),
  linesep = ""
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 1 presents the simulation results for the
two-timepoint case. Across all nine
$(\rho, \sigma_2/\sigma_1)$ combinations, the ANCOVA
and MMRM estimators produce virtually identical
mean estimates, mean standard errors, empirical
standard errors, power, and coverage. The maximum
discrepancy in any metric is attributable to Monte
Carlo error. This confirms the analytic result: with
complete data and an unstructured $2 \times 2$
covariance, MMRM reduces exactly to ANCOVA for the
treatment effect at follow-up.

```{r fig-scatter, fig.cap="Simulation-by-simulation comparison of treatment effect estimates from ANCOVA and MMRM. Each point represents one replication. Perfect agreement falls on the identity line.", fig.width=5, fig.height=4}
scatter_dat <- sim_results |>
  filter(
    rho == 0.6,
    sigma_ratio == 1.0,
    !is.na(est)
  ) |>
  select(sim, method, est) |>
  pivot_wider(names_from = method,
    values_from = est)

ggplot(scatter_dat, aes(x = ANCOVA, y = MMRM)) +
  geom_point(shape = 1, size = 0.8, alpha = 0.4) +
  geom_abline(
    intercept = 0, slope = 1,
    linetype = "dashed"
  ) +
  labs(
    x = "ANCOVA estimate",
    y = "MMRM estimate",
    title = expression(
      paste(rho, " = 0.6, ",
        sigma[2]/sigma[1], " = 1")
    )
  ) +
  theme_minimal(base_size = 10) +
  coord_equal()
```

```{r fig-se-comparison, fig.cap="Standard errors from ANCOVA versus MMRM across all nine scenarios. Points falling on the identity line indicate identical precision.", fig.width=5, fig.height=4}
se_dat <- sim_results |>
  filter(!is.na(se)) |>
  select(sim, method, se, rho, sigma_ratio) |>
  pivot_wider(names_from = method,
    values_from = se) |>
  mutate(
    scenario = paste0(
      "rho=", rho, ", ratio=", sigma_ratio
    )
  )

ggplot(se_dat, aes(x = ANCOVA, y = MMRM)) +
  geom_point(shape = 1, size = 0.5, alpha = 0.3) +
  geom_abline(
    intercept = 0, slope = 1,
    linetype = "dashed"
  ) +
  facet_wrap(~ scenario, scales = "free") +
  labs(
    x = "ANCOVA standard error",
    y = "MMRM standard error"
  ) +
  theme_minimal(base_size = 8)
```

\FloatBarrier

## Setting 2: three-timepoint random slopes

```{r sim-three-timepoint, cache=TRUE}
sim_three_tp <- function(n_per_arm, gamma_true,
                         sigma_b, sigma_e,
                         n_sims = 2000,
                         seed = 123) {
  set.seed(seed)
  results <- tibble(
    sim = integer(),
    method = character(),
    est = numeric(),
    se = numeric()
  )

  for (s in seq_len(n_sims)) {
    n <- 2 * n_per_arm
    g <- rep(0:1, each = n_per_arm)
    b_i <- rnorm(n, 0, sigma_b)
    t_vec <- c(0, 1, 2)

    dat <- expand_grid(
      id = seq_len(n),
      time = t_vec
    ) |>
      mutate(
        group = factor(g[id]),
        b = b_i[id],
        y = 50 + (2 + gamma_true *
          as.numeric(group == 1)) * time +
          b * time +
          rnorm(n(), 0, sigma_e)
      )

    c_dat <- dat |>
      arrange(id, time) |>
      group_by(id) |>
      mutate(
        change = y - lag(y)
      ) |>
      filter(!is.na(change)) |>
      ungroup()

    mean_change <- c_dat |>
      group_by(id) |>
      summarise(
        mean_c = mean(change),
        group = first(group),
        .groups = "drop"
      )

    fit_fixed <- lm(mean_c ~ group,
      data = mean_change)
    cf_f <- summary(fit_fixed)$coefficients
    est_f <- cf_f["group1", "Estimate"]
    se_f <- cf_f["group1", "Std. Error"]

    fit_re <- tryCatch({
      lme(y ~ time * group,
        random = ~ time | id,
        data = dat,
        control = lmeControl(
          opt = "optim",
          maxIter = 200
        ))
    }, error = function(e) NULL)

    if (!is.null(fit_re)) {
      cf_r <- summary(fit_re)$tTable
      est_r <- cf_r["time:group1", "Value"]
      se_r <- cf_r["time:group1", "Std.Error"]
    } else {
      est_r <- NA; se_r <- NA
    }

    results <- bind_rows(results, tibble(
      sim = s,
      method = c("Summary stat", "Random slopes"),
      est = c(est_f, est_r),
      se = c(se_f, se_r)
    ))
  }
  results
}

three_tp_scenarios <- expand_grid(
  sigma_b = c(0.5, 2, 8),
  sigma_e = c(4)
)

three_tp_results <- pmap_dfr(
  three_tp_scenarios,
  function(sigma_b, sigma_e) {
    res <- sim_three_tp(
      n_per_arm = 30,
      gamma_true = 2,
      sigma_b = sigma_b,
      sigma_e = sigma_e,
      n_sims = 2000
    )
    res |>
      mutate(
        sigma_b = sigma_b,
        sigma_e = sigma_e,
        ratio = sigma_b / sigma_e
      )
  }
)
```

```{r table-three-tp}
summary_3tp <- three_tp_results |>
  filter(!is.na(est)) |>
  group_by(sigma_b, sigma_e, ratio, method) |>
  summarise(
    mean_est = mean(est),
    mean_se = mean(se),
    emp_se = sd(est),
    .groups = "drop"
  ) |>
  mutate(
    re_ratio = round(
      mean_se[method == "Summary stat"] /
        mean_se[method == "Random slopes"],
      3
    ),
    .by = c(sigma_b, sigma_e)
  )

tab_3tp <- summary_3tp |>
  mutate(across(
    c(mean_est, mean_se, emp_se),
    ~ round(., 3)
  )) |>
  select(
    `$\\sigma_b$` = sigma_b,
    `$\\sigma_b/\\sigma$` = ratio,
    Method = method,
    `Mean $\\hat{\\gamma}$` = mean_est,
    `Mean SE` = mean_se,
    `Empirical SE` = emp_se,
    `SE ratio` = re_ratio
  )

kable(tab_3tp,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Comparison of summary-statistic and",
    "random-slopes estimators in three-timepoint",
    "designs ($n = 30$ per arm, $\\gamma = 2$,",
    "$\\sigma = 4$)."
  ),
  linesep = ""
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

When $\sigma_b / \sigma$ is small (0.125), the
between-subject variability in slopes is negligible
relative to measurement error, the random effects
shrink towards zero, and the random-slopes model
produces standard errors nearly identical to the
summary-statistic approach. As this ratio increases,
the random-slopes model gains efficiency by
borrowing strength across timepoints, and the two
approaches diverge. This confirms the
@frost2008optimizing observation: the critical ratio
is approximately $\sigma_b / \sigma = 4$
(equivalently, $2\sigma / \sigma_b = 1$ for equally
spaced designs), below which the simpler approach is
an adequate approximation.

```{r fig-ratio, fig.cap="Ratio of standard errors (summary statistic / random slopes) as a function of the between-subject slope SD to residual SD ratio. When the ratio is near 1, the two methods are equivalent.", fig.width=5, fig.height=3.5}
ratio_dat <- three_tp_results |>
  filter(!is.na(se)) |>
  select(sim, method, se, ratio) |>
  pivot_wider(names_from = method,
    values_from = se) |>
  mutate(
    se_ratio = `Summary stat` / `Random slopes`
  )

ratio_summary <- ratio_dat |>
  group_by(ratio) |>
  summarise(
    mean_ratio = mean(se_ratio, na.rm = TRUE),
    lo = quantile(se_ratio, 0.025, na.rm = TRUE),
    hi = quantile(se_ratio, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ratio_summary,
  aes(x = ratio, y = mean_ratio)) +
  geom_point(shape = 16, size = 3) +
  geom_errorbar(
    aes(ymin = lo, ymax = hi),
    width = 0.02
  ) +
  geom_hline(
    yintercept = 1,
    linetype = "dashed", colour = "grey50"
  ) +
  scale_x_continuous(
    breaks = c(0.125, 0.5, 2)
  ) +
  labs(
    x = expression(sigma[b] / sigma),
    y = "SE ratio (summary stat / random slopes)"
  ) +
  theme_minimal(base_size = 10)
```

\FloatBarrier

# Discussion

The results establish three conditions under which
random effects models reduce to or closely
approximate fixed-effect alternatives.

**Condition 1: unstructured covariance with two
timepoints.** When MMRM is fit with an unrestricted
$2 \times 2$ covariance matrix, the treatment
effect estimate and its standard error are
algebraically identical to those from ANCOVA. This
is not an approximation but an exact equivalence,
implicit in @lee1974note and explicit in
@frost2008optimizing (Section 2.2). The practical
implication is that for any two-timepoint design
with complete data, ANCOVA is a sufficient analysis
and the MMRM apparatus adds no information about
the treatment contrast.

**Condition 2: small between-subject slope variance.**
In designs with three or more timepoints, the
random-slopes model becomes equivalent to the
summary-statistic analysis when $\sigma_b / \sigma$
is small. As @frost2008optimizing showed, when this
ratio is below approximately 0.25 (for equally
spaced designs with the same total follow-up),
the efficiency gain from the random effects model
is negligible. In practical terms, this corresponds
to settings where measurement error dominates
biological variability in rates---a situation common
in short trials with noisy endpoints.

**Condition 3: cLDA with complete data.** The
constrained longitudinal model produces treatment
effect estimates effectively identical to ANCOVA
when no baseline values are missing
[@liu2009baseline; @lu2010efficiency]. The marginal
advantage of cLDA emerges only with incomplete
baselines, where it gains 1--4\% additional power
depending on the missingness rate.

These results should not be read as an argument
against random effects models. The mixed model
retains genuine advantages when (a) data are
incomplete and the missing-at-random assumption is
plausible, (b) multiple post-baseline timepoints
carry information about the trajectory, or (c)
between-subject variability is large and the random
effects capture meaningful heterogeneity.
@sullivan2022mmrm demonstrated that MMRM with
time-by-covariance interactions dominates ANCOVA in
the presence of dropout, and @wang2024improving
showed that augmented MMRM can further improve
robustness.

The practical value of these equivalence results is
pedagogical and communicative. When an investigator
asks what the MMRM ``is doing,'' the analyst can
respond: in your two-visit design, it is doing
exactly what ANCOVA does, no more and no less. The
random effects are not contributing additional
information; they are merely a different
parameterization of the same model. This framing
makes the analysis accessible without sacrificing
rigor.

## Limitations

The simulations assumed balanced designs with equal
group sizes and no missing data (except in the cLDA
comparison). The equivalence results extend
straightforwardly to unbalanced designs but may be
perturbed by informative missingness, small samples
where the Kenward-Roger correction
[@kenward1997small] matters, or misspecification of
the covariance structure. We did not investigate
designs with more than two treatment groups, where
the mapping between MMRM contrasts and ANCOVA
contrasts is more complex.

# Future research

1. **Extension to multiple post-baseline visits.**
   The exact equivalence in the two-timepoint case
   does not hold with three or more visits unless
   the covariance is unstructured. Characterizing
   the discrepancy as a function of the number of
   visits and the covariance structure is of
   practical interest.

2. **Non-normal outcomes.** The relationship between
   generalized linear mixed models and marginal
   (GEE) models for binary and count endpoints
   involves additional complexities due to the
   non-collapsibility of odds ratios and rate ratios.

3. **Informative missingness.** Under
   missing-not-at-random mechanisms, the equivalence
   between MMRM and ANCOVA breaks down in different
   ways. Sensitivity analyses comparing the two
   frameworks under various departure patterns from
   MAR would be informative.

4. **Software tools.** An R package providing
   automatic detection of when MMRM reduces to a
   simpler model---based on the number of timepoints,
   completeness, and estimated variance-component
   ratios---would be useful for practitioners.

5. **Pedagogical materials.** Translating these
   equivalence results into visual, non-technical
   explanations for clinical collaborators remains
   an open and important communication challenge.

# References
