---
title: "Donna B Project. A comparison of MMRM and LM for simple designs"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
geometry: "left=3cm,right=3in,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
    includes:
      in_header: "~/shr/preamble.tex"
header-includes:
   - \rhead{Lab report}
   - \fancyfoot[L]{\currfilename} %put current file in footer
   - \setlength{\headheight}{13.59999pt}
---

<!-- your comment 
\input{notes.tex}
\hspace{4in}
\fbox{
\parbox{3in}{\Large 
Notes: 
\begin{itemize}
\item a \\
\item b  \\
\end{itemize}
}
}
-->

```{r  include=FALSE, echo=F}
rm(list = ls())
options(
  knitr.kable.NA = "", dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9, knitr.table.format = "latex"
)
library(pacman)
p_load(skimr, janitor, kableExtra, tidyverse, knitr, lubridate, readxl)

opts_chunk$set(
  warning = F, message = F, echo = F, fig.width = 3.2,
  fig.height = 3, results = "asis", dev = "pdf",
  opts_chunk$set(cache.path = "cache/")
)
source("~/shr/zz.tools.R")
options(scipen = 1, digits = 2)
```

<!-- your comment 
\listoftodos
\lipsum[2]
\todo[inline, color=cyan]{The original todo note withouth changed colours.\newline Here's another line.}
\begin{wrapfigure}{r}[.4\textwidth]{.4\textwidth}%
\textit{\sffamily\normalsize\textcolor{gray!50}{
Tornado plot: x-axis participants IDs y-axis study visits plot mean of cumulative
AEs stratified by pre-covid and post-covid study visits (March 12 is
the date we are using as a cut-point: through March 12 is pre-COVID,
after March 12 is post-COVID) }}% 
\end{wrapfigure}
\lipsum[2]

Sample reference section reference. @Pocock1975a

\begin{marginfigure}
\includegraphics[width=3in]{/Users/zenn/shr/images/sudoku-3d-cube.pdf} \caption{test}\label{fig:unnamed-chunk-8}
\end{marginfigure}

My most common usage of the todonotes package, is to insert an uncustomized
todonotes somewhere in a latex document. An example of this usage is the command
\todo{Make a cake \ldots},
\lipsum[1]
\todo[color=green]{Make a pie \ldots},
\lipsum[1] 
\todo{Make a joke \ldots},
\listoftodos
-->



# Introduction

The question arises as to whether in the case of two treatment groups and two
timepoints does MMRM reduce exactly to ANCOVA.

* Does if matter if the analysis is adjusting for baseline (i.e. not including
  baseline in MMRM response.).

* Does it matter if response is observed value or is a change score?
Consider the article by Brogan and Kutner [@Brogan1980]

\FloatBarrier

# Methods

Set up the model, no covariates,  in both settings with 3 observations per
group. Work out analytic formula

* Repeat for unequal variances

* Repeat w covariates

# Methods

# Results
# Code 

<!-- your comment 
```{r, attr.source='.numberLines', ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
-->

# References

\bibliography{x24.bib}
\bibliographystyle{plain}
\nocite{*}

